---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Week 6 results

## Useful links

### Video

### Other sources

+ MABS Book chapter ~ https://www.crcpress.com/Multivariate-Analysis-for-the-Behavioral-Sciences-Second-Edition/Vehkalahti-Everitt/p/book/9780815385158

### Data

https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt

https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt

## Task 1: Implement the analyses of Chapter 8 of MABS using the RATS data 

### Load data
```{r}
RATS = read.csv("./ratsl.csv")
str(RATS)

RATS$Group = factor(RATS$Group)
RATS$ID = factor(RATS$ID)
```

### Scale data
```{r}
#Access the package ggplot2
library(ggplot2)

# Draw the plot
ggplot(RATS, aes(x = wd, y = rats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$rats), max(RATS$rats)))

# Standardise the variable bprs
RATS <- RATS %>%
  group_by(wd) %>%
  mutate(stdrats = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS)

# Plot again with the standardised rats
ggplot(RATS, aes(x = wd, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "standardized rats", limits = c(min(RATS$stdrats), max(RATS$stdrats)))


```

### Summarize data 

```{r}

# Number of wd, baseline (wd1) included
n <- RATS$wd %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATS %>%
  group_by(Group, wd) %>%
  summarise( mean = mean(rats), se = sd(rats)/sqrt(n)) %>%
               ungroup()
             
# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = wd, y = mean, linetype = Group, shape = Group)) +
 geom_line() +
 scale_linetype_manual(values = c(1,2,3)) +
 geom_point(size=3) +
 scale_shape_manual(values = c(1,2,3)) +
 geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
 theme(legend.position = c(0.8,0.5)) +
 scale_y_continuous(name = "mean(rats) +/- se(rats)")

```

### Pick data without the first week

```{r}

RATS8S <- RATS %>%
  filter(wd > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATS8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), wd 1-65")
```

### Filter out outliers

```{r}

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATS8S1 <- RATS %>%
  filter((wd > 1)&((rats > 250 & rats < 550))) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS8S1)

# Draw a boxplot of the mean versus treatment
ggplot(RATS8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), wd 1-65")

```

### Anova

```{r}
ratsOrig <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep  ="\t", header = T)
str(ratsOrig)

```



```{r}

# Perform a two-sample t-test
#t.test(mean ~ Group, data = RATS8S1, var.equal = TRUE)

# Add the baseline from the original data as a new variable to the summary data
RATS8S2 <- RATS8S %>%
  mutate(baseline = ratsOrig$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATS8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

### Conclusions

The Anova results imply that the is not enough evidence that the different groups would develop differently in the treatment when the baseline is cancelled from the observation results, but on risk level 10% it would be so.

## Task 2

### Load data (already in long format)
```{r}

BPRSL = read.csv("./bprsl.csv")
str(BPRSL)

BPRSL$treatment = factor(BPRSL$treatment)
BPRSL$subject = factor(BPRSL$subject)

# Glimpse the data

glimpse(BPRSL)
```

### Data plot
```{r}
# Check the dimensions of the data
dim(BPRSL)

# Plot the data
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

### Regression model

```{r}
# create a regression model
BPRS_reg <- lm(bprs ~ treatment + week, data = BPRSL)
# print out a summary of the model
summary(BPRS_reg)

```

### Random intercept model

```{r}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

Note: Pay attention to variability (standard deviation) of the subject

### Random intercept and random slope model

```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

Note: Pay attention to the chi-squared statistics and p-value of the likelihood ratio test between BPRS_ref1 and BPRS_ref. The lower the value the better the fit against the comparison model.

### Create a random intercept and random slope model with the interaction

```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + week*treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)


```

### Model fit, graphical interpretation

```{r}
# draw the plot of BPRS with the observed bprs values
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to BPRSL
 BPRSL <- mutate(BPRSL, fit = Fitted)

# draw the plot of BPRS with the Fitted values of bprs
ggplot(BPRSL, aes(x = week, y = fit, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

